---
title: "Modelling YF vaccination coverage (1970 to 2025 update)"
output: html_document
date: "2026-04-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Packages

```{r packages, message=FALSE, warning=FALSE}
library(tidyr)
library(ggplot2)
library(dplyr)
library(scales)
library(sf)
library(readr)
library(stringr)
library (grid) #for map cropping
library (ggrepel) #for labeling maps
```

## Replace missing values with 0

```{r clamp}
clamp01 <- function(x) { x[is.na(x)] <- 0; pmin(pmax(x, 0), 1) }
```

## Model as a function (one district at a time)

Notes:
-This model currently uses baseline vaccination coverage in 1970 from Shearer (optimistic) vaccination coverage found in https://pubmed.ncbi.nlm.nih.gov/28822780/.
-To change the source for baseline coverage, change all instances of "shearer_opt" to "shearer_un" (targeted), "shearer_con" (conservative) or "polici" (Polici data found in https://polici.shinyapps.io/yellow_fever_africa/).
-This model has no routine vaccination data for the years 1992, 1993, 1994, 1995, 1996, 1997, 1998 and 1999.
-This model includes all supplementary vaccination campaigns from the years 2005, 2006, 2012, 2018, 2020, 2021 and 2022. It is missing data from 2011 and 2012 campaigns (44 districts targeted, covering 5.8 million people).

```{r yfmodel}
run_model_one_district <- function(df) {
  
  df <- df %>%
    filter(year >= 1970) %>%
    mutate(year = as.numeric(year)) %>%
    arrange(year)
  
  #Stop early if there's not enough data
  if (nrow(df) < 2) {
    return(df[0, ])
  }
  
  #Checking required/missing columns
  required_cols <- c("year","infants","total_pop","shearer_opt","routine","supplementary")
  missing_cols <- setdiff(required_cols, names(df))
  if (length(missing_cols) > 0) {
    stop("Missing required columns for district ", unique(df$district),
         ": ", paste(missing_cols, collapse = ", "))
  }
  
  #Turn columns into numeric format
  num_cols <- c("infants","total_pop","shearer_opt","routine","supplementary")
  df[num_cols] <- lapply(df[num_cols], function(x) as.numeric(as.character(x)))
  
  #Error out any N/As/check if there are negatives
  if (anyNA(df$infants) || anyNA(df$total_pop)) {
    bad <- which(is.na(df$infants) | is.na(df$total_pop))[1]
    stop("NA in infants/total_pop at district ", unique(df$district),
         ", year ", df$year[bad], ". Fix the CSV.")
  }
  if (any(df$infants < 0, na.rm = TRUE) || any(df$total_pop < 0, na.rm = TRUE)) {
    stop("Negative infants or total_pop found in district ", unique(df$district), ". Fix the CSV.")
  }
  if (any(df$infants > df$total_pop, na.rm = TRUE)) {
    bad <- which(df$infants > df$total_pop)[1]
    stop("infants > total_pop at district ", unique(df$district),
         ", year ", df$year[bad], ". Fix the CSV.")
  }
  
  #Ensure that coverage/percentage is between 0 and 1; treat NA as 0
  df$shearer_opt    <- clamp01(df$shearer_opt)
  df$routine        <- clamp01(df$routine)
  df$supplementary  <- clamp01(df$supplementary)
  
  #Allocate storage columns (states and flows)
  df <- df %>%
    mutate(
      #Number of people in each state
      U_inf = NA_real_, V_inf = NA_real_,
      U_noninf = NA_real_, V_noninf = NA_real_,
      #Totals
      vaccinated_total = NA_real_,
      Coverage_stock = NA_real_,
      RI_flow = NA_real_,
      #Flows (results of each year)
      SIA_flow_inf = NA_real_,
      SIA_flow_noninf = NA_real_,
      SIA_flow = NA_real_,
      Total_flow = NA_real_
    )
  
  #Baseline year (t = 1)
  t1 <- 1
  infants1    <- df$infants[t1]
  total_pop1  <- df$total_pop[t1]
  noninf_pop1 <- total_pop1 - infants1
  
  U_inf_prev    <- infants1
  V_inf_prev    <- 0
  V_noninf_prev <- df$shearer_opt[t1] * noninf_pop1
  U_noninf_prev <- noninf_pop1 - V_noninf_prev
  
  df$U_inf[t1]    <- U_inf_prev
  df$V_inf[t1]    <- V_inf_prev
  df$U_noninf[t1] <- U_noninf_prev
  df$V_noninf[t1] <- V_noninf_prev
  
  df$vaccinated_total[t1] <- V_inf_prev + V_noninf_prev
  df$Coverage_stock[t1]   <- df$vaccinated_total[t1] / total_pop1
  
  df$RI_flow[t1]         <- 0
  df$SIA_flow_inf[t1]    <- 0
  df$SIA_flow_noninf[t1] <- 0
  df$SIA_flow[t1]        <- 0
  df$Total_flow[t1]      <- 0
  
  #Simulation loop
  for (t in 2:nrow(df)) {
    
    infants    <- df$infants[t]
    total_pop  <- df$total_pop[t]
    noninf_pop <- total_pop - infants
    
    r <- df$routine[t]
    s <- df$supplementary[t]
    
    #1) New infants enter
    U_inf <- infants
    V_inf <- 0
    
    #2) Routine immunization among infants only
    RI <- r * U_inf
    U_inf_after_RI <- U_inf - RI
    V_inf_after_RI <- V_inf + RI
    
    #3) Age last year's infants into non-infants
    U_noninf_temp <- U_noninf_prev + U_inf_prev
    V_noninf_temp <- V_noninf_prev + V_inf_prev
    
    #4) SIA applies to unvaccinated (infants + noninfants) only
    total_unvax_before_SIA <- max(U_inf_after_RI + U_noninf_temp, 0)
    
    SIA_total <- min(s * total_unvax_before_SIA, total_unvax_before_SIA)
    
    if (total_unvax_before_SIA > 0) {
      SIA_inf    <- SIA_total * (U_inf_after_RI / total_unvax_before_SIA)
      SIA_noninf <- SIA_total - SIA_inf
    } else {
      SIA_inf <- 0
      SIA_noninf <- 0
    }
    
    #Apply transfers (moving people from unvaccinated to vaccinated)
    U_inf_final    <- max(U_inf_after_RI - SIA_inf, 0)
    V_inf_final    <- max(V_inf_after_RI + SIA_inf, 0)
    
    U_noninf_final <- max(U_noninf_temp - SIA_noninf, 0)
    V_noninf_final <- max(V_noninf_temp + SIA_noninf, 0)
    
    #5) Mass balance via proportional rescaling to match total population
    inf_total <- U_inf_final + V_inf_final
    if (inf_total > 0) {
      scale_inf <- infants / inf_total
      U_inf_final <- U_inf_final * scale_inf
      V_inf_final <- V_inf_final * scale_inf
    } else {
      U_inf_final <- infants
      V_inf_final <- 0
    }
    
    noninf_total <- U_noninf_final + V_noninf_final
    if (noninf_total > 0) {
      scale_noninf <- noninf_pop / noninf_total
      U_noninf_final <- U_noninf_final * scale_noninf
      V_noninf_final <- V_noninf_final * scale_noninf
    } else {
      U_noninf_final <- noninf_pop
      V_noninf_final <- 0
    }
    
    #Store results for each year
    df$U_inf[t]    <- U_inf_final
    df$V_inf[t]    <- V_inf_final
    df$U_noninf[t] <- U_noninf_final
    df$V_noninf[t] <- V_noninf_final
    
    df$vaccinated_total[t] <- V_inf_final + V_noninf_final
    df$Coverage_stock[t]   <- if (total_pop > 0) df$vaccinated_total[t] / total_pop else NA_real_
    
    #Store flows
    df$RI_flow[t]         <- RI
    df$SIA_flow_inf[t]    <- SIA_inf
    df$SIA_flow_noninf[t] <- SIA_noninf
    df$SIA_flow[t]        <- SIA_total
    df$Total_flow[t]      <- RI + SIA_total
    
    #Carry forward (how many vaccinations happened this year))
    U_inf_prev    <- U_inf_final
    V_inf_prev    <- V_inf_final
    U_noninf_prev <- U_noninf_final
    V_noninf_prev <- V_noninf_final
  }
  
  #Check that total number of infants/non-infants makes sense
  stopifnot(all(abs((df$U_inf + df$V_inf) - df$infants) < 1e-6))
  stopifnot(all(abs((df$U_noninf + df$V_noninf) - (df$total_pop - df$infants)) < 1e-6))
  
  df
}
```

## Run model for all districts

```{r allmodel}
raw <- read.csv("optimistic1970.csv")

results <- raw %>%
  group_by(district) %>%
  group_modify(~ run_model_one_district(.x)) %>%
  ungroup()
```

## Write coverage stock to .csv (for map)

This command produces the "coverage_by_district_year1970.csv" file, which contains estimated vaccination coverages per district per year from 1970 - 2025.

```{r coveragestock}
results %>%
  select(district, year, Coverage_stock) %>%
  write.csv("coverage_by_district_year_opt1970.csv", row.names = FALSE)
```

## Organize data sets for mapping

```{r mapprep}

#Shapefile (.dbf and .shx files should be in same folder)
districts_sf <- st_read("261_Districts.shp", quiet = FALSE)

#Coverage data
cov <- read_csv("coverage_by_district_year_opt1970.csv",
                col_types = cols(
                  district = col_character(),
                  year = col_integer(),
                  Coverage_stock = col_double()
                )) %>%
  filter(year >= 1970, year <= 2025) 

#Clean district names to match both datasets
clean_name <- function(x) {
  x %>%
    str_to_lower() %>%
    str_trim() %>%
    str_replace_all("[[:punct:]]", " ") %>%
    str_replace_all("\\s+", " ")
}

districts_sf2 <- districts_sf %>%
  mutate(
    district_key = clean_name(District)  # <-- replace DIST_NAME_FIELD
  )

cov2 <- cov %>%
  mutate(
    district_key = clean_name(district)
  )

#Check if district names match
unmatched_cov <- cov2 %>%
  anti_join(districts_sf2 %>% st_drop_geometry() %>% select(district_key),
            by = "district_key") %>%
  distinct(district) %>%
  arrange(district)

unmatched_shp <- districts_sf2 %>%
  st_drop_geometry() %>%
  anti_join(cov2 %>% select(district_key) %>% distinct(),
            by = "district_key") %>%
  distinct(district_key)

unmatched_cov
unmatched_shp

#Join geometry x coverage (missing years become N/A)
years <- 1970:2025

# Full grid: every district in shapefile for every year
grid <- expand_grid(
  district_key = districts_sf2 %>% st_drop_geometry() %>% pull(district_key) %>% unique(),
  year = years
)

cov_full <- grid %>%
  left_join(cov2 %>% select(district_key, year, Coverage_stock),
            by = c("district_key", "year"))

map_data <- districts_sf2 %>%
  left_join(cov_full, by = "district_key")
```

##Make vaccination coverage map

```{r facetmap}
p <- ggplot(map_data) +
  geom_sf(aes(fill = Coverage_stock),
          color = "black",
          linewidth = 0.10) +
  facet_wrap(~ year, ncol = 5) +
  scale_fill_stepsn(
    colours = c("#8c1d18", "#d7301f", "#fc8d59", "#FFD166",
                "#e5f5f9", "#99d8c9", "#2ca25f"),
    breaks  = c(0, 0.5, 0.6, 0.7, 0.8, 0.9, 0.95, 1.0),
    limits  = c(0, 1),
    labels = function(x) {
      labs <- percent(x, accuracy = 1)
      show <- x %in% c(0, 0.6, 0.8, 0.9, 1.0)  # key thresholds
      labs[!show] <- ""
      labs
    },
    na.value = "grey85",
    name = "Vaccination coverage"
  ) +
  labs(
    title = "Yellow Fever vaccination coverage by district (Ghana), 1970–2025",
    subtitle = "(Optimistic)"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    strip.text = element_text(size = 10),
    legend.position = "bottom",
    legend.key.width = unit(30, "pt"),
    legend.title = element_text(margin = margin(b = 5)),
    legend.text = element_text(size = 9)
  ) +
  guides(
    fill = guide_colorsteps(
      title.position = "top",
      label.position = "bottom",
      barwidth = unit(14, "cm"),
      barheight = unit(0.6, "cm")
    )
  )

p
print(p)
ggsave("ghana_coverage_facets_1970_2025opt.png", p, width = 16, height = 10, dpi = 300)
```

```{r}
p <- ggplot(map_data) +
  geom_sf(aes(fill = Coverage_stock),
          color = "black",
          linewidth = 0.10) +
  coord_sf(expand = FALSE) +
  facet_wrap(~ year, ncol = 6) +
  scale_fill_stepsn(
    colours = c("#8c1d18", "#d7301f", "#fc8d59", "#FFD166",
                "#e5f5f9", "#99d8c9", "#2ca25f"),
    breaks  = c(0, 0.5, 0.6, 0.7, 0.8, 0.9, 0.95, 1.0),
    limits  = c(0, 1),
    labels = function(x) {
      labs <- percent(x, accuracy = 1)
      show <- x %in% c(0, 0.6, 0.8, 0.9, 1.0)  # key thresholds
      labs[!show] <- ""
      labs
    },
    na.value = "grey85",
    name = "Vaccination coverage"
  ) +
  labs(
    title = "Yellow Fever vaccination coverage by district (Ghana), 1970–2025",
    subtitle = "(Optimistic)"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    strip.text = element_text(size = 10),
    legend.position = "bottom",
    legend.key.width = unit(30, "pt"),
    legend.title = element_text(margin = margin(b = 5)),
    legend.text = element_text(size = 9)
  ) +
  theme(
  panel.spacing = unit(1, "lines")
) +
  guides(
    fill = guide_colorsteps(
      title.position = "top",
      label.position = "bottom",
      barwidth = unit(14, "cm"),
      barheight = unit(0.6, "cm")
    )
  )

p
print(p)
ggsave(
  "ghana_coverage_facets_1970_2025opt_portrait.png",
  p,
  width = 14,   
  height = 24,  
  dpi = 300
)
```
```{r}
p <- ggplot(map_data) +
  geom_sf(aes(fill = Coverage_stock),
          color = "black",
          linewidth = 0.10) +
  coord_sf(expand = FALSE) +
  facet_wrap(~ year, ncol = 12) +
  scale_fill_stepsn(
    colours = c("#8c1d18", "#d7301f", "#fc8d59", "#FFD166",
                "#e5f5f9", "#99d8c9", "#2ca25f"),
    breaks  = c(0, 0.5, 0.6, 0.7, 0.8, 0.9, 0.95, 1.0),
    limits  = c(0, 1),
    labels = function(x) {
      labs <- percent(x, accuracy = 1)
      show <- x %in% c(0, 0.6, 0.8, 0.9, 1.0)  # key thresholds
      labs[!show] <- ""
      labs
    },
    na.value = "grey85",
    name = "Vaccination coverage"
  ) +
  labs(
    title = "Yellow Fever vaccination coverage by district (Ghana), 1970–2025",
    subtitle = "(Optimistic)"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    strip.text = element_text(size = 10),
    legend.position = "bottom",
    legend.key.width = unit(30, "pt"),
    legend.title = element_text(margin = margin(b = 5)),
    legend.text = element_text(size = 9)
  ) +
  theme(
  panel.spacing = unit(1, "lines")
) +
  guides(
    fill = guide_colorsteps(
      title.position = "top",
      label.position = "bottom",
      barwidth = unit(14, "cm"),
      barheight = unit(0.6, "cm")
    )
  )

p
print(p)
ggsave(
  "ghana_coverage_facets_1970_2025opt_landscape.png",
  p,
  width = 24,   
  height = 14,  
  dpi = 300
)
```

## This uses conservative Shearer data. Running the model in this first chunk actually isn't necessary as shearer_un, shearer_con and shearer_opt coverages are all the same in 1970.

```{r}
run_model_one_district <- function(df) {
  
  df <- df %>%
    filter(year >= 1970) %>%
    mutate(year = as.numeric(year)) %>%
    arrange(year)
  
  #Stop early if there's not enough data
  if (nrow(df) < 2) {
    return(df[0, ])
  }
  
  #Checking required/missing columns
  required_cols <- c("year","infants","total_pop","shearer_con","routine","supplementary")
  missing_cols <- setdiff(required_cols, names(df))
  if (length(missing_cols) > 0) {
    stop("Missing required columns for district ", unique(df$district),
         ": ", paste(missing_cols, collapse = ", "))
  }
  
  #Turn columns into numeric format
  num_cols <- c("infants","total_pop","shearer_con","routine","supplementary")
  df[num_cols] <- lapply(df[num_cols], function(x) as.numeric(as.character(x)))
  
  #Error out any N/As/check if there are negatives
  if (anyNA(df$infants) || anyNA(df$total_pop)) {
    bad <- which(is.na(df$infants) | is.na(df$total_pop))[1]
    stop("NA in infants/total_pop at district ", unique(df$district),
         ", year ", df$year[bad], ". Fix the CSV.")
  }
  if (any(df$infants < 0, na.rm = TRUE) || any(df$total_pop < 0, na.rm = TRUE)) {
    stop("Negative infants or total_pop found in district ", unique(df$district), ". Fix the CSV.")
  }
  if (any(df$infants > df$total_pop, na.rm = TRUE)) {
    bad <- which(df$infants > df$total_pop)[1]
    stop("infants > total_pop at district ", unique(df$district),
         ", year ", df$year[bad], ". Fix the CSV.")
  }
  
  #Ensure that coverage/percentage is between 0 and 1; treat NA as 0
  df$shearer_con    <- clamp01(df$shearer_con)
  df$routine        <- clamp01(df$routine)
  df$supplementary  <- clamp01(df$supplementary)
  
  #Allocate storage columns (states and flows)
  df <- df %>%
    mutate(
      #Number of people in each state
      U_inf = NA_real_, V_inf = NA_real_,
      U_noninf = NA_real_, V_noninf = NA_real_,
      #Totals
      vaccinated_total = NA_real_,
      Coverage_stock = NA_real_,
      RI_flow = NA_real_,
      #Flows (results of each year)
      SIA_flow_inf = NA_real_,
      SIA_flow_noninf = NA_real_,
      SIA_flow = NA_real_,
      Total_flow = NA_real_
    )
  
  #Baseline year (t = 1)
  t1 <- 1
  infants1    <- df$infants[t1]
  total_pop1  <- df$total_pop[t1]
  noninf_pop1 <- total_pop1 - infants1
  
  U_inf_prev    <- infants1
  V_inf_prev    <- 0
  V_noninf_prev <- df$shearer_con[t1] * noninf_pop1
  U_noninf_prev <- noninf_pop1 - V_noninf_prev
  
  df$U_inf[t1]    <- U_inf_prev
  df$V_inf[t1]    <- V_inf_prev
  df$U_noninf[t1] <- U_noninf_prev
  df$V_noninf[t1] <- V_noninf_prev
  
  df$vaccinated_total[t1] <- V_inf_prev + V_noninf_prev
  df$Coverage_stock[t1]   <- df$vaccinated_total[t1] / total_pop1
  
  df$RI_flow[t1]         <- 0
  df$SIA_flow_inf[t1]    <- 0
  df$SIA_flow_noninf[t1] <- 0
  df$SIA_flow[t1]        <- 0
  df$Total_flow[t1]      <- 0
  
  #Simulation loop
  for (t in 2:nrow(df)) {
    
    infants    <- df$infants[t]
    total_pop  <- df$total_pop[t]
    noninf_pop <- total_pop - infants
    
    r <- df$routine[t]
    s <- df$supplementary[t]
    
    #1) New infants enter
    U_inf <- infants
    V_inf <- 0
    
    #2) Routine immunization among infants only
    RI <- r * U_inf
    U_inf_after_RI <- U_inf - RI
    V_inf_after_RI <- V_inf + RI
    
    #3) Age last year's infants into non-infants
    U_noninf_temp <- U_noninf_prev + U_inf_prev
    V_noninf_temp <- V_noninf_prev + V_inf_prev
    
    #4) SIA applies to unvaccinated (infants + noninfants) only
    total_unvax_before_SIA <- max(U_inf_after_RI + U_noninf_temp, 0)
    
    SIA_total <- min(s * total_unvax_before_SIA, total_unvax_before_SIA)
    
    if (total_unvax_before_SIA > 0) {
      SIA_inf    <- SIA_total * (U_inf_after_RI / total_unvax_before_SIA)
      SIA_noninf <- SIA_total - SIA_inf
    } else {
      SIA_inf <- 0
      SIA_noninf <- 0
    }
    
    #Apply transfers (moving people from unvaccinated to vaccinated)
    U_inf_final    <- max(U_inf_after_RI - SIA_inf, 0)
    V_inf_final    <- max(V_inf_after_RI + SIA_inf, 0)
    
    U_noninf_final <- max(U_noninf_temp - SIA_noninf, 0)
    V_noninf_final <- max(V_noninf_temp + SIA_noninf, 0)
    
    #5) Mass balance via proportional rescaling to match total population
    inf_total <- U_inf_final + V_inf_final
    if (inf_total > 0) {
      scale_inf <- infants / inf_total
      U_inf_final <- U_inf_final * scale_inf
      V_inf_final <- V_inf_final * scale_inf
    } else {
      U_inf_final <- infants
      V_inf_final <- 0
    }
    
    noninf_total <- U_noninf_final + V_noninf_final
    if (noninf_total > 0) {
      scale_noninf <- noninf_pop / noninf_total
      U_noninf_final <- U_noninf_final * scale_noninf
      V_noninf_final <- V_noninf_final * scale_noninf
    } else {
      U_noninf_final <- noninf_pop
      V_noninf_final <- 0
    }
    
    #Store results for each year
    df$U_inf[t]    <- U_inf_final
    df$V_inf[t]    <- V_inf_final
    df$U_noninf[t] <- U_noninf_final
    df$V_noninf[t] <- V_noninf_final
    
    df$vaccinated_total[t] <- V_inf_final + V_noninf_final
    df$Coverage_stock[t]   <- if (total_pop > 0) df$vaccinated_total[t] / total_pop else NA_real_
    
    #Store flows
    df$RI_flow[t]         <- RI
    df$SIA_flow_inf[t]    <- SIA_inf
    df$SIA_flow_noninf[t] <- SIA_noninf
    df$SIA_flow[t]        <- SIA_total
    df$Total_flow[t]      <- RI + SIA_total
    
    #Carry forward (how many vaccinations happened this year))
    U_inf_prev    <- U_inf_final
    V_inf_prev    <- V_inf_final
    U_noninf_prev <- U_noninf_final
    V_noninf_prev <- V_noninf_final
  }
  
  #Check that total number of infants/non-infants makes sense
  stopifnot(all(abs((df$U_inf + df$V_inf) - df$infants) < 1e-6))
  stopifnot(all(abs((df$U_noninf + df$V_noninf) - (df$total_pop - df$infants)) < 1e-6))
  
  df
}
```

```{r}
raw <- read.csv("conservative1970.csv")

results <- raw %>%
  group_by(district) %>%
  group_modify(~ run_model_one_district(.x)) %>%
  ungroup()
```


```{r}
results %>%
  select(district, year, Coverage_stock) %>%
  write.csv("coverage_by_district_year_con1970.csv", row.names = FALSE)
```

```{r}

#Shapefile (.dbf and .shx files should be in same folder)
districts_sf <- st_read("261_Districts.shp", quiet = FALSE)

#Coverage data
cov <- read_csv("coverage_by_district_year_con1970.csv",
                col_types = cols(
                  district = col_character(),
                  year = col_integer(),
                  Coverage_stock = col_double()
                )) %>%
  filter(year >= 1970, year <= 2025) 

#Clean district names to match both datasets
clean_name <- function(x) {
  x %>%
    str_to_lower() %>%
    str_trim() %>%
    str_replace_all("[[:punct:]]", " ") %>%
    str_replace_all("\\s+", " ")
}

districts_sf2 <- districts_sf %>%
  mutate(
    district_key = clean_name(District)  # <-- replace DIST_NAME_FIELD
  )

cov2 <- cov %>%
  mutate(
    district_key = clean_name(district)
  )

#Check if district names match
unmatched_cov <- cov2 %>%
  anti_join(districts_sf2 %>% st_drop_geometry() %>% select(district_key),
            by = "district_key") %>%
  distinct(district) %>%
  arrange(district)

unmatched_shp <- districts_sf2 %>%
  st_drop_geometry() %>%
  anti_join(cov2 %>% select(district_key) %>% distinct(),
            by = "district_key") %>%
  distinct(district_key)

unmatched_cov
unmatched_shp

#Join geometry x coverage (missing years become N/A)
years <- 1970:2025

# Full grid: every district in shapefile for every year
grid <- expand_grid(
  district_key = districts_sf2 %>% st_drop_geometry() %>% pull(district_key) %>% unique(),
  year = years
)

cov_full <- grid %>%
  left_join(cov2 %>% select(district_key, year, Coverage_stock),
            by = c("district_key", "year"))

map_data <- districts_sf2 %>%
  left_join(cov_full, by = "district_key")
```


```{r}
p <- ggplot(map_data) +
  geom_sf(aes(fill = Coverage_stock),
          color = "black",
          linewidth = 0.10) +
  coord_sf(expand = FALSE) +
  facet_wrap(~ year, ncol = 6) +
  scale_fill_stepsn(
    colours = c("#8c1d18", "#d7301f", "#fc8d59", "#FFD166",
                "#e5f5f9", "#99d8c9", "#2ca25f"),
    breaks  = c(0, 0.5, 0.6, 0.7, 0.8, 0.9, 0.95, 1.0),
    limits  = c(0, 1),
    labels = function(x) {
      labs <- percent(x, accuracy = 1)
      show <- x %in% c(0, 0.6, 0.8, 0.9, 1.0)  # key thresholds
      labs[!show] <- ""
      labs
    },
    na.value = "grey85",
    name = "Vaccination coverage"
  ) +
  labs(
    title = "Yellow Fever vaccination coverage by district (Ghana), 1970–2025",
    subtitle = "(Conservative)"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    strip.text = element_text(size = 10),
    legend.position = "bottom",
    legend.key.width = unit(30, "pt"),
    legend.title = element_text(margin = margin(b = 5)),
    legend.text = element_text(size = 9)
  ) +
  theme(
  panel.spacing = unit(1, "lines")
) +
  guides(
    fill = guide_colorsteps(
      title.position = "top",
      label.position = "bottom",
      barwidth = unit(14, "cm"),
      barheight = unit(0.6, "cm")
    )
  )

p
print(p)
ggsave(
  "ghana_coverage_facets_1970_2025con_portrait.png",
  p,
  width = 14,   
  height = 24,  
  dpi = 300
)
```

```{r}
p <- ggplot(map_data) +
  geom_sf(aes(fill = Coverage_stock),
          color = "black",
          linewidth = 0.10) +
  coord_sf(expand = FALSE) +
  facet_wrap(~ year, ncol = 12) +
  scale_fill_stepsn(
    colours = c("#8c1d18", "#d7301f", "#fc8d59", "#FFD166",
                "#e5f5f9", "#99d8c9", "#2ca25f"),
    breaks  = c(0, 0.5, 0.6, 0.7, 0.8, 0.9, 0.95, 1.0),
    limits  = c(0, 1),
    labels = function(x) {
      labs <- percent(x, accuracy = 1)
      show <- x %in% c(0, 0.6, 0.8, 0.9, 1.0)  # key thresholds
      labs[!show] <- ""
      labs
    },
    na.value = "grey85",
    name = "Vaccination coverage"
  ) +
  labs(
    title = "Yellow Fever vaccination coverage by district (Ghana), 1970–2025",
    subtitle = "(Conservative)"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    strip.text = element_text(size = 10),
    legend.position = "bottom",
    legend.key.width = unit(30, "pt"),
    legend.title = element_text(margin = margin(b = 5)),
    legend.text = element_text(size = 9)
  ) +
  theme(
  panel.spacing = unit(1, "lines")
) +
  guides(
    fill = guide_colorsteps(
      title.position = "top",
      label.position = "bottom",
      barwidth = unit(14, "cm"),
      barheight = unit(0.6, "cm")
    )
  )

p
print(p)
ggsave(
  "ghana_coverage_facets_1970_2025con_landscape.png",
  p,
  width = 24,   
  height = 14,  
  dpi = 300
)
```

## This one uses Shearer untargeted/unbiased coverage. Running the model in this first chunk actually isn't necessary as shearer_un, shearer_con and shearer_opt coverages are all the same in 1970.

```{r}
run_model_one_district <- function(df) {
  
  df <- df %>%
    filter(year >= 1970) %>%
    mutate(year = as.numeric(year)) %>%
    arrange(year)
  
  #Stop early if there's not enough data
  if (nrow(df) < 2) {
    return(df[0, ])
  }
  
  #Checking required/missing columns
  required_cols <- c("year","infants","total_pop","shearer_un","routine","supplementary")
  missing_cols <- setdiff(required_cols, names(df))
  if (length(missing_cols) > 0) {
    stop("Missing required columns for district ", unique(df$district),
         ": ", paste(missing_cols, collapse = ", "))
  }
  
  #Turn columns into numeric format
  num_cols <- c("infants","total_pop","shearer_un","routine","supplementary")
  df[num_cols] <- lapply(df[num_cols], function(x) as.numeric(as.character(x)))
  
  #Error out any N/As/check if there are negatives
  if (anyNA(df$infants) || anyNA(df$total_pop)) {
    bad <- which(is.na(df$infants) | is.na(df$total_pop))[1]
    stop("NA in infants/total_pop at district ", unique(df$district),
         ", year ", df$year[bad], ". Fix the CSV.")
  }
  if (any(df$infants < 0, na.rm = TRUE) || any(df$total_pop < 0, na.rm = TRUE)) {
    stop("Negative infants or total_pop found in district ", unique(df$district), ". Fix the CSV.")
  }
  if (any(df$infants > df$total_pop, na.rm = TRUE)) {
    bad <- which(df$infants > df$total_pop)[1]
    stop("infants > total_pop at district ", unique(df$district),
         ", year ", df$year[bad], ". Fix the CSV.")
  }
  
  #Ensure that coverage/percentage is between 0 and 1; treat NA as 0
  df$shearer_un    <- clamp01(df$shearer_un)
  df$routine        <- clamp01(df$routine)
  df$supplementary  <- clamp01(df$supplementary)
  
  #Allocate storage columns (states and flows)
  df <- df %>%
    mutate(
      #Number of people in each state
      U_inf = NA_real_, V_inf = NA_real_,
      U_noninf = NA_real_, V_noninf = NA_real_,
      #Totals
      vaccinated_total = NA_real_,
      Coverage_stock = NA_real_,
      RI_flow = NA_real_,
      #Flows (results of each year)
      SIA_flow_inf = NA_real_,
      SIA_flow_noninf = NA_real_,
      SIA_flow = NA_real_,
      Total_flow = NA_real_
    )
  
  #Baseline year (t = 1)
  t1 <- 1
  infants1    <- df$infants[t1]
  total_pop1  <- df$total_pop[t1]
  noninf_pop1 <- total_pop1 - infants1
  
  U_inf_prev    <- infants1
  V_inf_prev    <- 0
  V_noninf_prev <- df$shearer_un[t1] * noninf_pop1
  U_noninf_prev <- noninf_pop1 - V_noninf_prev
  
  df$U_inf[t1]    <- U_inf_prev
  df$V_inf[t1]    <- V_inf_prev
  df$U_noninf[t1] <- U_noninf_prev
  df$V_noninf[t1] <- V_noninf_prev
  
  df$vaccinated_total[t1] <- V_inf_prev + V_noninf_prev
  df$Coverage_stock[t1]   <- df$vaccinated_total[t1] / total_pop1
  
  df$RI_flow[t1]         <- 0
  df$SIA_flow_inf[t1]    <- 0
  df$SIA_flow_noninf[t1] <- 0
  df$SIA_flow[t1]        <- 0
  df$Total_flow[t1]      <- 0
  
  #Simulation loop
  for (t in 2:nrow(df)) {
    
    infants    <- df$infants[t]
    total_pop  <- df$total_pop[t]
    noninf_pop <- total_pop - infants
    
    r <- df$routine[t]
    s <- df$supplementary[t]
    
    #1) New infants enter
    U_inf <- infants
    V_inf <- 0
    
    #2) Routine immunization among infants only
    RI <- r * U_inf
    U_inf_after_RI <- U_inf - RI
    V_inf_after_RI <- V_inf + RI
    
    #3) Age last year's infants into non-infants
    U_noninf_temp <- U_noninf_prev + U_inf_prev
    V_noninf_temp <- V_noninf_prev + V_inf_prev
    
    #4) SIA applies to unvaccinated (infants + noninfants) only
    total_unvax_before_SIA <- max(U_inf_after_RI + U_noninf_temp, 0)
    
    SIA_total <- min(s * total_unvax_before_SIA, total_unvax_before_SIA)
    
    if (total_unvax_before_SIA > 0) {
      SIA_inf    <- SIA_total * (U_inf_after_RI / total_unvax_before_SIA)
      SIA_noninf <- SIA_total - SIA_inf
    } else {
      SIA_inf <- 0
      SIA_noninf <- 0
    }
    
    #Apply transfers (moving people from unvaccinated to vaccinated)
    U_inf_final    <- max(U_inf_after_RI - SIA_inf, 0)
    V_inf_final    <- max(V_inf_after_RI + SIA_inf, 0)
    
    U_noninf_final <- max(U_noninf_temp - SIA_noninf, 0)
    V_noninf_final <- max(V_noninf_temp + SIA_noninf, 0)
    
    #5) Mass balance via proportional rescaling to match total population
    inf_total <- U_inf_final + V_inf_final
    if (inf_total > 0) {
      scale_inf <- infants / inf_total
      U_inf_final <- U_inf_final * scale_inf
      V_inf_final <- V_inf_final * scale_inf
    } else {
      U_inf_final <- infants
      V_inf_final <- 0
    }
    
    noninf_total <- U_noninf_final + V_noninf_final
    if (noninf_total > 0) {
      scale_noninf <- noninf_pop / noninf_total
      U_noninf_final <- U_noninf_final * scale_noninf
      V_noninf_final <- V_noninf_final * scale_noninf
    } else {
      U_noninf_final <- noninf_pop
      V_noninf_final <- 0
    }
    
    #Store results for each year
    df$U_inf[t]    <- U_inf_final
    df$V_inf[t]    <- V_inf_final
    df$U_noninf[t] <- U_noninf_final
    df$V_noninf[t] <- V_noninf_final
    
    df$vaccinated_total[t] <- V_inf_final + V_noninf_final
    df$Coverage_stock[t]   <- if (total_pop > 0) df$vaccinated_total[t] / total_pop else NA_real_
    
    #Store flows
    df$RI_flow[t]         <- RI
    df$SIA_flow_inf[t]    <- SIA_inf
    df$SIA_flow_noninf[t] <- SIA_noninf
    df$SIA_flow[t]        <- SIA_total
    df$Total_flow[t]      <- RI + SIA_total
    
    #Carry forward (how many vaccinations happened this year))
    U_inf_prev    <- U_inf_final
    V_inf_prev    <- V_inf_final
    U_noninf_prev <- U_noninf_final
    V_noninf_prev <- V_noninf_final
  }
  
  #Check that total number of infants/non-infants makes sense
  stopifnot(all(abs((df$U_inf + df$V_inf) - df$infants) < 1e-6))
  stopifnot(all(abs((df$U_noninf + df$V_noninf) - (df$total_pop - df$infants)) < 1e-6))
  
  df
}
```

```{r}
raw <- read.csv("untargeted1970.csv")

results <- raw %>%
  group_by(district) %>%
  group_modify(~ run_model_one_district(.x)) %>%
  ungroup()
```

```{r}
results %>%
  select(district, year, Coverage_stock) %>%
  write.csv("coverage_by_district_year_un1970.csv", row.names = FALSE)
```

```{r}

#Shapefile (.dbf and .shx files should be in same folder)
districts_sf <- st_read("261_Districts.shp", quiet = FALSE)

#Coverage data
cov <- read_csv("coverage_by_district_year_un1970.csv",
                col_types = cols(
                  district = col_character(),
                  year = col_integer(),
                  Coverage_stock = col_double()
                )) %>%
  filter(year >= 1970, year <= 2025) 

#Clean district names to match both datasets
clean_name <- function(x) {
  x %>%
    str_to_lower() %>%
    str_trim() %>%
    str_replace_all("[[:punct:]]", " ") %>%
    str_replace_all("\\s+", " ")
}

districts_sf2 <- districts_sf %>%
  mutate(
    district_key = clean_name(District)  # <-- replace DIST_NAME_FIELD
  )

cov2 <- cov %>%
  mutate(
    district_key = clean_name(district)
  )

#Check if district names match
unmatched_cov <- cov2 %>%
  anti_join(districts_sf2 %>% st_drop_geometry() %>% select(district_key),
            by = "district_key") %>%
  distinct(district) %>%
  arrange(district)

unmatched_shp <- districts_sf2 %>%
  st_drop_geometry() %>%
  anti_join(cov2 %>% select(district_key) %>% distinct(),
            by = "district_key") %>%
  distinct(district_key)

unmatched_cov
unmatched_shp

#Join geometry x coverage (missing years become N/A)
years <- 1970:2025

# Full grid: every district in shapefile for every year
grid <- expand_grid(
  district_key = districts_sf2 %>% st_drop_geometry() %>% pull(district_key) %>% unique(),
  year = years
)

cov_full <- grid %>%
  left_join(cov2 %>% select(district_key, year, Coverage_stock),
            by = c("district_key", "year"))

map_data <- districts_sf2 %>%
  left_join(cov_full, by = "district_key")
```


```{r}
p <- ggplot(map_data) +
  geom_sf(aes(fill = Coverage_stock),
          color = "black",
          linewidth = 0.10) +
  coord_sf(expand = FALSE) +
  facet_wrap(~ year, ncol = 6) +
  scale_fill_stepsn(
    colours = c("#8c1d18", "#d7301f", "#fc8d59", "#FFD166",
                "#e5f5f9", "#99d8c9", "#2ca25f"),
    breaks  = c(0, 0.5, 0.6, 0.7, 0.8, 0.9, 0.95, 1.0),
    limits  = c(0, 1),
    labels = function(x) {
      labs <- percent(x, accuracy = 1)
      show <- x %in% c(0, 0.6, 0.8, 0.9, 1.0)  # key thresholds
      labs[!show] <- ""
      labs
    },
    na.value = "grey85",
    name = "Vaccination coverage"
  ) +
  labs(
    title = "Yellow Fever vaccination coverage by district (Ghana), 1970–2025",
    subtitle = "(Untargeted)"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    strip.text = element_text(size = 10),
    legend.position = "bottom",
    legend.key.width = unit(30, "pt"),
    legend.title = element_text(margin = margin(b = 5)),
    legend.text = element_text(size = 9)
  ) +
  theme(
  panel.spacing = unit(1, "lines")
) +
  guides(
    fill = guide_colorsteps(
      title.position = "top",
      label.position = "bottom",
      barwidth = unit(14, "cm"),
      barheight = unit(0.6, "cm")
    )
  )

p
print(p)
ggsave(
  "ghana_coverage_facets_1970_2025un_portrait.png",
  p,
  width = 14,   
  height = 24,  
  dpi = 300
)
```

```{r}
p <- ggplot(map_data) +
  geom_sf(aes(fill = Coverage_stock),
          color = "black",
          linewidth = 0.10) +
  coord_sf(expand = FALSE) +
  facet_wrap(~ year, ncol = 12) +
  scale_fill_stepsn(
    colours = c("#8c1d18", "#d7301f", "#fc8d59", "#FFD166",
                "#e5f5f9", "#99d8c9", "#2ca25f"),
    breaks  = c(0, 0.5, 0.6, 0.7, 0.8, 0.9, 0.95, 1.0),
    limits  = c(0, 1),
    labels = function(x) {
      labs <- percent(x, accuracy = 1)
      show <- x %in% c(0, 0.6, 0.8, 0.9, 1.0)  # key thresholds
      labs[!show] <- ""
      labs
    },
    na.value = "grey85",
    name = "Vaccination coverage"
  ) +
  labs(
    title = "Yellow Fever vaccination coverage by district (Ghana), 1970–2025",
    subtitle = "(Untargeted)"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    strip.text = element_text(size = 10),
    legend.position = "bottom",
    legend.key.width = unit(30, "pt"),
    legend.title = element_text(margin = margin(b = 5)),
    legend.text = element_text(size = 9)
  ) +
  theme(
  panel.spacing = unit(1, "lines")
) +
  guides(
    fill = guide_colorsteps(
      title.position = "top",
      label.position = "bottom",
      barwidth = unit(14, "cm"),
      barheight = unit(0.6, "cm")
    )
  )

p
print(p)
ggsave(
  "ghana_coverage_facets_1970_2025un_landscape.png",
  p,
  width = 24,   
  height = 14,  
  dpi = 300
)
```

## This code simulates 100% vaccination coverage in 2011-2012 in certain districts due to the supplementary mass vaccination campaigns in those years. We still don't have precise coverage data, so vaccination coverage has been set to 100% for most districts instead.

```{r}
run_model_one_district <- function(df) {
  
  df <- df %>%
    filter(year >= 1970) %>%
    mutate(year = as.numeric(year)) %>%
    arrange(year)
  
  #Stop early if there's not enough data
  if (nrow(df) < 2) {
    return(df[0, ])
  }
  
  #Checking required/missing columns
  required_cols <- c("year","infants","total_pop","shearer_opt","routine","supplementary")
  missing_cols <- setdiff(required_cols, names(df))
  if (length(missing_cols) > 0) {
    stop("Missing required columns for district ", unique(df$district),
         ": ", paste(missing_cols, collapse = ", "))
  }
  
  #Turn columns into numeric format
  num_cols <- c("infants","total_pop","shearer_opt","routine","supplementary")
  df[num_cols] <- lapply(df[num_cols], function(x) as.numeric(as.character(x)))
  
  #Error out any N/As/check if there are negatives
  if (anyNA(df$infants) || anyNA(df$total_pop)) {
    bad <- which(is.na(df$infants) | is.na(df$total_pop))[1]
    stop("NA in infants/total_pop at district ", unique(df$district),
         ", year ", df$year[bad], ". Fix the CSV.")
  }
  if (any(df$infants < 0, na.rm = TRUE) || any(df$total_pop < 0, na.rm = TRUE)) {
    stop("Negative infants or total_pop found in district ", unique(df$district), ". Fix the CSV.")
  }
  if (any(df$infants > df$total_pop, na.rm = TRUE)) {
    bad <- which(df$infants > df$total_pop)[1]
    stop("infants > total_pop at district ", unique(df$district),
         ", year ", df$year[bad], ". Fix the CSV.")
  }
  
  #Ensure that coverage/percentage is between 0 and 1; treat NA as 0
  df$shearer_opt    <- clamp01(df$shearer_opt)
  df$routine        <- clamp01(df$routine)
  df$supplementary  <- clamp01(df$supplementary)
  
  #Allocate storage columns (states and flows)
  df <- df %>%
    mutate(
      #Number of people in each state
      U_inf = NA_real_, V_inf = NA_real_,
      U_noninf = NA_real_, V_noninf = NA_real_,
      #Totals
      vaccinated_total = NA_real_,
      Coverage_stock = NA_real_,
      RI_flow = NA_real_,
      #Flows (results of each year)
      SIA_flow_inf = NA_real_,
      SIA_flow_noninf = NA_real_,
      SIA_flow = NA_real_,
      Total_flow = NA_real_
    )
  
  #Baseline year (t = 1)
  t1 <- 1
  infants1    <- df$infants[t1]
  total_pop1  <- df$total_pop[t1]
  noninf_pop1 <- total_pop1 - infants1
  
  U_inf_prev    <- infants1
  V_inf_prev    <- 0
  V_noninf_prev <- df$shearer_opt[t1] * noninf_pop1
  U_noninf_prev <- noninf_pop1 - V_noninf_prev
  
  df$U_inf[t1]    <- U_inf_prev
  df$V_inf[t1]    <- V_inf_prev
  df$U_noninf[t1] <- U_noninf_prev
  df$V_noninf[t1] <- V_noninf_prev
  
  df$vaccinated_total[t1] <- V_inf_prev + V_noninf_prev
  df$Coverage_stock[t1]   <- df$vaccinated_total[t1] / total_pop1
  
  df$RI_flow[t1]         <- 0
  df$SIA_flow_inf[t1]    <- 0
  df$SIA_flow_noninf[t1] <- 0
  df$SIA_flow[t1]        <- 0
  df$Total_flow[t1]      <- 0
  
  #Simulation loop
  for (t in 2:nrow(df)) {
    
    infants    <- df$infants[t]
    total_pop  <- df$total_pop[t]
    noninf_pop <- total_pop - infants
    
    r <- df$routine[t]
    s <- df$supplementary[t]
    
    #1) New infants enter
    U_inf <- infants
    V_inf <- 0
    
    #2) Routine immunization among infants only
    RI <- r * U_inf
    U_inf_after_RI <- U_inf - RI
    V_inf_after_RI <- V_inf + RI
    
    #3) Age last year's infants into non-infants
    U_noninf_temp <- U_noninf_prev + U_inf_prev
    V_noninf_temp <- V_noninf_prev + V_inf_prev
    
    #4) SIA applies to unvaccinated (infants + noninfants) only
    total_unvax_before_SIA <- max(U_inf_after_RI + U_noninf_temp, 0)
    
    SIA_total <- min(s * total_unvax_before_SIA, total_unvax_before_SIA)
    
    if (total_unvax_before_SIA > 0) {
      SIA_inf    <- SIA_total * (U_inf_after_RI / total_unvax_before_SIA)
      SIA_noninf <- SIA_total - SIA_inf
    } else {
      SIA_inf <- 0
      SIA_noninf <- 0
    }
    
    #Apply transfers (moving people from unvaccinated to vaccinated)
    U_inf_final    <- max(U_inf_after_RI - SIA_inf, 0)
    V_inf_final    <- max(V_inf_after_RI + SIA_inf, 0)
    
    U_noninf_final <- max(U_noninf_temp - SIA_noninf, 0)
    V_noninf_final <- max(V_noninf_temp + SIA_noninf, 0)
    
    #5) Mass balance via proportional rescaling to match total population
    inf_total <- U_inf_final + V_inf_final
    if (inf_total > 0) {
      scale_inf <- infants / inf_total
      U_inf_final <- U_inf_final * scale_inf
      V_inf_final <- V_inf_final * scale_inf
    } else {
      U_inf_final <- infants
      V_inf_final <- 0
    }
    
    noninf_total <- U_noninf_final + V_noninf_final
    if (noninf_total > 0) {
      scale_noninf <- noninf_pop / noninf_total
      U_noninf_final <- U_noninf_final * scale_noninf
      V_noninf_final <- V_noninf_final * scale_noninf
    } else {
      U_noninf_final <- noninf_pop
      V_noninf_final <- 0
    }
    
    #Store results for each year
    df$U_inf[t]    <- U_inf_final
    df$V_inf[t]    <- V_inf_final
    df$U_noninf[t] <- U_noninf_final
    df$V_noninf[t] <- V_noninf_final
    
    df$vaccinated_total[t] <- V_inf_final + V_noninf_final
    df$Coverage_stock[t]   <- if (total_pop > 0) df$vaccinated_total[t] / total_pop else NA_real_
    
    #Store flows
    df$RI_flow[t]         <- RI
    df$SIA_flow_inf[t]    <- SIA_inf
    df$SIA_flow_noninf[t] <- SIA_noninf
    df$SIA_flow[t]        <- SIA_total
    df$Total_flow[t]      <- RI + SIA_total
    
    #Carry forward (how many vaccinations happened this year))
    U_inf_prev    <- U_inf_final
    V_inf_prev    <- V_inf_final
    U_noninf_prev <- U_noninf_final
    V_noninf_prev <- V_noninf_final
  }
  
  #Check that total number of infants/non-infants makes sense
  stopifnot(all(abs((df$U_inf + df$V_inf) - df$infants) < 1e-6))
  stopifnot(all(abs((df$U_noninf + df$V_noninf) - (df$total_pop - df$infants)) < 1e-6))
  
  df
}
```

## Run model for all districts

```{r}
raw <- read.csv("optimistic1970_supplementary2011.csv")

results <- raw %>%
  group_by(district) %>%
  group_modify(~ run_model_one_district(.x)) %>%
  ungroup()
```

## Table for district-level coverage

```{r}
results %>%
  select(district, year, Coverage_stock) %>%
  write.csv("coverage_by_district_year_opt1970_s2011.csv", row.names = FALSE)
```

## Organize data sets for mapping

```{r}

#Shapefile (.dbf and .shx files should be in same folder)
districts_sf <- st_read("261_Districts.shp", quiet = FALSE)

#Coverage data
cov <- read_csv("coverage_by_district_year_opt1970_s2011.csv",
                col_types = cols(
                  district = col_character(),
                  year = col_integer(),
                  Coverage_stock = col_double()
                )) %>%
  filter(year >= 1970, year <= 2025) 

#Clean district names to match both datasets
clean_name <- function(x) {
  x %>%
    str_to_lower() %>%
    str_trim() %>%
    str_replace_all("[[:punct:]]", " ") %>%
    str_replace_all("\\s+", " ")
}

districts_sf2 <- districts_sf %>%
  mutate(
    district_key = clean_name(District)  # <-- replace DIST_NAME_FIELD
  )

cov2 <- cov %>%
  mutate(
    district_key = clean_name(district)
  )

#Check if district names match
unmatched_cov <- cov2 %>%
  anti_join(districts_sf2 %>% st_drop_geometry() %>% select(district_key),
            by = "district_key") %>%
  distinct(district) %>%
  arrange(district)

unmatched_shp <- districts_sf2 %>%
  st_drop_geometry() %>%
  anti_join(cov2 %>% select(district_key) %>% distinct(),
            by = "district_key") %>%
  distinct(district_key)

unmatched_cov
unmatched_shp

#Join geometry x coverage (missing years become N/A)
years <- 1970:2025

# Full grid: every district in shapefile for every year
grid <- expand_grid(
  district_key = districts_sf2 %>% st_drop_geometry() %>% pull(district_key) %>% unique(),
  year = years
)

cov_full <- grid %>%
  left_join(cov2 %>% select(district_key, year, Coverage_stock),
            by = c("district_key", "year"))

map_data <- districts_sf2 %>%
  left_join(cov_full, by = "district_key")
```

```{r}
p <- ggplot(map_data) +
  geom_sf(aes(fill = Coverage_stock),
          color = "black",
          linewidth = 0.10) +
  coord_sf(expand = FALSE) +
  facet_wrap(~ year, ncol = 6) +
  scale_fill_stepsn(
    colours = c("#8c1d18", "#d7301f", "#fc8d59", "#FFD166",
                "#e5f5f9", "#99d8c9", "#2ca25f"),
    breaks  = c(0, 0.5, 0.6, 0.7, 0.8, 0.9, 0.95, 1.0),
    limits  = c(0, 1),
    labels = function(x) {
      labs <- percent(x, accuracy = 1)
      show <- x %in% c(0, 0.6, 0.8, 0.9, 1.0)  # key thresholds
      labs[!show] <- ""
      labs
    },
    na.value = "grey85",
    name = "Vaccination coverage"
  ) +
  labs(
    title = "Yellow Fever vaccination coverage by district (Ghana), 1970–2025",
    subtitle = "(Includes 2011-2012 campaigns)"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    strip.text = element_text(size = 10),
    legend.position = "bottom",
    legend.key.width = unit(30, "pt"),
    legend.title = element_text(margin = margin(b = 5)),
    legend.text = element_text(size = 9)
  ) +
  theme(
  panel.spacing = unit(1, "lines")
) +
  guides(
    fill = guide_colorsteps(
      title.position = "top",
      label.position = "bottom",
      barwidth = unit(14, "cm"),
      barheight = unit(0.6, "cm")
    )
  )

p
print(p)
ggsave(
  "ghana_coverage_facets_1970_2025_2011s_portrait.png",
  p,
  width = 14,   
  height = 24,  
  dpi = 300
)
```

```{r}
p <- ggplot(map_data) +
  geom_sf(aes(fill = Coverage_stock),
          color = "black",
          linewidth = 0.10) +
  coord_sf(expand = FALSE) +
  facet_wrap(~ year, ncol = 12) +
  scale_fill_stepsn(
    colours = c("#8c1d18", "#d7301f", "#fc8d59", "#FFD166",
                "#e5f5f9", "#99d8c9", "#2ca25f"),
    breaks  = c(0, 0.5, 0.6, 0.7, 0.8, 0.9, 0.95, 1.0),
    limits  = c(0, 1),
    labels = function(x) {
      labs <- percent(x, accuracy = 1)
      show <- x %in% c(0, 0.6, 0.8, 0.9, 1.0)  # key thresholds
      labs[!show] <- ""
      labs
    },
    na.value = "grey85",
    name = "Vaccination coverage"
  ) +
  labs(
    title = "Yellow Fever vaccination coverage by district (Ghana), 1970–2025",
    subtitle = "(Includes 2011-2012 campaigns)"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    strip.text = element_text(size = 10),
    legend.position = "bottom",
    legend.key.width = unit(30, "pt"),
    legend.title = element_text(margin = margin(b = 5)),
    legend.text = element_text(size = 9)
  ) +
  theme(
  panel.spacing = unit(1, "lines")
) +
  guides(
    fill = guide_colorsteps(
      title.position = "top",
      label.position = "bottom",
      barwidth = unit(14, "cm"),
      barheight = unit(0.6, "cm")
    )
  )

p
print(p)
ggsave(
  "ghana_coverage_facets_1970_2025_2011s_landscape.png",
  p,
  width = 24,   
  height = 14,  
  dpi = 300
)
```

##This code is for producing a 2005-2025 map only.

```{r}
run_model_one_district <- function(df) {
  
  df <- df %>%
    filter(year >= 2000) %>%
    mutate(year = as.numeric(year)) %>%
    arrange(year)
  
  #Stop early if there's not enough data
  if (nrow(df) < 2) {
    return(df[0, ])
  }
  
  #Checking required/missing columns
  required_cols <- c("year","infants","total_pop","shearer_opt","routine","supplementary")
  missing_cols <- setdiff(required_cols, names(df))
  if (length(missing_cols) > 0) {
    stop("Missing required columns for district ", unique(df$district),
         ": ", paste(missing_cols, collapse = ", "))
  }
  
  #Turn columns into numeric format
  num_cols <- c("infants","total_pop","shearer_opt","routine","supplementary")
  df[num_cols] <- lapply(df[num_cols], function(x) as.numeric(as.character(x)))
  
  #Error out any N/As/check if there are negatives
  if (anyNA(df$infants) || anyNA(df$total_pop)) {
    bad <- which(is.na(df$infants) | is.na(df$total_pop))[1]
    stop("NA in infants/total_pop at district ", unique(df$district),
         ", year ", df$year[bad], ". Fix the CSV.")
  }
  if (any(df$infants < 0, na.rm = TRUE) || any(df$total_pop < 0, na.rm = TRUE)) {
    stop("Negative infants or total_pop found in district ", unique(df$district), ". Fix the CSV.")
  }
  if (any(df$infants > df$total_pop, na.rm = TRUE)) {
    bad <- which(df$infants > df$total_pop)[1]
    stop("infants > total_pop at district ", unique(df$district),
         ", year ", df$year[bad], ". Fix the CSV.")
  }
  
  #Ensure that coverage/percentage is between 0 and 1; treat NA as 0
  df$shearer_opt    <- clamp01(df$shearer_opt)
  df$routine        <- clamp01(df$routine)
  df$supplementary  <- clamp01(df$supplementary)
  
  #Allocate storage columns (states and flows)
  df <- df %>%
    mutate(
      #Number of people in each state
      U_inf = NA_real_, V_inf = NA_real_,
      U_noninf = NA_real_, V_noninf = NA_real_,
      #Totals
      vaccinated_total = NA_real_,
      Coverage_stock = NA_real_,
      RI_flow = NA_real_,
      #Flows (results of each year)
      SIA_flow_inf = NA_real_,
      SIA_flow_noninf = NA_real_,
      SIA_flow = NA_real_,
      Total_flow = NA_real_
    )
  
  #Baseline year (t = 1)
  t1 <- 1
  infants1    <- df$infants[t1]
  total_pop1  <- df$total_pop[t1]
  noninf_pop1 <- total_pop1 - infants1
  
  U_inf_prev    <- infants1
  V_inf_prev    <- 0
  V_noninf_prev <- df$shearer_opt[t1] * noninf_pop1
  U_noninf_prev <- noninf_pop1 - V_noninf_prev
  
  df$U_inf[t1]    <- U_inf_prev
  df$V_inf[t1]    <- V_inf_prev
  df$U_noninf[t1] <- U_noninf_prev
  df$V_noninf[t1] <- V_noninf_prev
  
  df$vaccinated_total[t1] <- V_inf_prev + V_noninf_prev
  df$Coverage_stock[t1]   <- df$vaccinated_total[t1] / total_pop1
  
  df$RI_flow[t1]         <- 0
  df$SIA_flow_inf[t1]    <- 0
  df$SIA_flow_noninf[t1] <- 0
  df$SIA_flow[t1]        <- 0
  df$Total_flow[t1]      <- 0
  
  #Simulation loop
  for (t in 2:nrow(df)) {
    
    infants    <- df$infants[t]
    total_pop  <- df$total_pop[t]
    noninf_pop <- total_pop - infants
    
    r <- df$routine[t]
    s <- df$supplementary[t]
    
    #1) New infants enter
    U_inf <- infants
    V_inf <- 0
    
    #2) Routine immunization among infants only
    RI <- r * U_inf
    U_inf_after_RI <- U_inf - RI
    V_inf_after_RI <- V_inf + RI
    
    #3) Age last year's infants into non-infants
    U_noninf_temp <- U_noninf_prev + U_inf_prev
    V_noninf_temp <- V_noninf_prev + V_inf_prev
    
    #4) SIA applies to unvaccinated (infants + noninfants) only
    total_unvax_before_SIA <- max(U_inf_after_RI + U_noninf_temp, 0)
    
    SIA_total <- min(s * total_unvax_before_SIA, total_unvax_before_SIA)
    
    if (total_unvax_before_SIA > 0) {
      SIA_inf    <- SIA_total * (U_inf_after_RI / total_unvax_before_SIA)
      SIA_noninf <- SIA_total - SIA_inf
    } else {
      SIA_inf <- 0
      SIA_noninf <- 0
    }
    
    #Apply transfers (moving people from unvaccinated to vaccinated)
    U_inf_final    <- max(U_inf_after_RI - SIA_inf, 0)
    V_inf_final    <- max(V_inf_after_RI + SIA_inf, 0)
    
    U_noninf_final <- max(U_noninf_temp - SIA_noninf, 0)
    V_noninf_final <- max(V_noninf_temp + SIA_noninf, 0)
    
    #5) Mass balance via proportional rescaling to match total population
    inf_total <- U_inf_final + V_inf_final
    if (inf_total > 0) {
      scale_inf <- infants / inf_total
      U_inf_final <- U_inf_final * scale_inf
      V_inf_final <- V_inf_final * scale_inf
    } else {
      U_inf_final <- infants
      V_inf_final <- 0
    }
    
    noninf_total <- U_noninf_final + V_noninf_final
    if (noninf_total > 0) {
      scale_noninf <- noninf_pop / noninf_total
      U_noninf_final <- U_noninf_final * scale_noninf
      V_noninf_final <- V_noninf_final * scale_noninf
    } else {
      U_noninf_final <- noninf_pop
      V_noninf_final <- 0
    }
    
    #Store results for each year
    df$U_inf[t]    <- U_inf_final
    df$V_inf[t]    <- V_inf_final
    df$U_noninf[t] <- U_noninf_final
    df$V_noninf[t] <- V_noninf_final
    
    df$vaccinated_total[t] <- V_inf_final + V_noninf_final
    df$Coverage_stock[t]   <- if (total_pop > 0) df$vaccinated_total[t] / total_pop else NA_real_
    
    #Store flows
    df$RI_flow[t]         <- RI
    df$SIA_flow_inf[t]    <- SIA_inf
    df$SIA_flow_noninf[t] <- SIA_noninf
    df$SIA_flow[t]        <- SIA_total
    df$Total_flow[t]      <- RI + SIA_total
    
    #Carry forward (how many vaccinations happened this year))
    U_inf_prev    <- U_inf_final
    V_inf_prev    <- V_inf_final
    U_noninf_prev <- U_noninf_final
    V_noninf_prev <- V_noninf_final
  }
  
  #Check that total number of infants/non-infants makes sense
  stopifnot(all(abs((df$U_inf + df$V_inf) - df$infants) < 1e-6))
  stopifnot(all(abs((df$U_noninf + df$V_noninf) - (df$total_pop - df$infants)) < 1e-6))
  
  df
}
```

## Run model for all districts

```{r}
raw <- read.csv("optimistic1970.csv")

results <- raw %>%
  group_by(district) %>%
  group_modify(~ run_model_one_district(.x)) %>%
  ungroup()
```

## Write coverage stock to .csv (for map)

This command produces the "coverage_by_district_year1970.csv" file, which contains estimated vaccination coverages per district per year from 1970 - 2025.

```{r}
results %>%
  select(district, year, Coverage_stock) %>%
  write.csv("coverage_by_district_year_opt2000.csv", row.names = FALSE)
```

## Organize data sets for mapping

```{r}

#Shapefile (.dbf and .shx files should be in same folder)
districts_sf <- st_read("261_Districts.shp", quiet = FALSE)

#Coverage data
cov <- read_csv("coverage_by_district_year_opt2000.csv",
                col_types = cols(
                  district = col_character(),
                  year = col_integer(),
                  Coverage_stock = col_double()
                )) %>%
  filter(year >= 2005, year <= 2025) 

#Clean district names to match both datasets
clean_name <- function(x) {
  x %>%
    str_to_lower() %>%
    str_trim() %>%
    str_replace_all("[[:punct:]]", " ") %>%
    str_replace_all("\\s+", " ")
}

districts_sf2 <- districts_sf %>%
  mutate(
    district_key = clean_name(District)  # <-- replace DIST_NAME_FIELD
  )

cov2 <- cov %>%
  mutate(
    district_key = clean_name(district)
  )

#Check if district names match
unmatched_cov <- cov2 %>%
  anti_join(districts_sf2 %>% st_drop_geometry() %>% select(district_key),
            by = "district_key") %>%
  distinct(district) %>%
  arrange(district)

unmatched_shp <- districts_sf2 %>%
  st_drop_geometry() %>%
  anti_join(cov2 %>% select(district_key) %>% distinct(),
            by = "district_key") %>%
  distinct(district_key)

unmatched_cov
unmatched_shp

#Join geometry x coverage (missing years become N/A)
years <- 2005:2025

# Full grid: every district in shapefile for every year
grid <- expand_grid(
  district_key = districts_sf2 %>% st_drop_geometry() %>% pull(district_key) %>% unique(),
  year = years
)

cov_full <- grid %>%
  left_join(cov2 %>% select(district_key, year, Coverage_stock),
            by = c("district_key", "year"))

map_data <- districts_sf2 %>%
  left_join(cov_full, by = "district_key")
```

```{r}
p <- ggplot(map_data) +
  geom_sf(aes(fill = Coverage_stock),
          color = "black",
          linewidth = 0.10) +
  coord_sf(expand = FALSE) +
  facet_wrap(~ year, ncol = 5) +
  scale_fill_stepsn(
    colours = c("#8c1d18", "#d7301f", "#fc8d59", "#FFD166",
                "#e5f5f9", "#99d8c9", "#2ca25f"),
    breaks  = c(0, 0.5, 0.6, 0.7, 0.8, 0.9, 0.95, 1.0),
    limits  = c(0, 1),
    labels = function(x) {
      labs <- percent(x, accuracy = 1)
      show <- x %in% c(0, 0.6, 0.8, 0.9, 1.0)  # key thresholds
      labs[!show] <- ""
      labs
    },
    na.value = "grey85",
    name = "Vaccination coverage"
  ) +
  labs(
    title = "Yellow Fever vaccination coverage by district (Ghana), 2005–2025" #,
    #subtitle = "(Optimistic)"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    strip.text = element_text(size = 10),
    legend.position = "bottom",
    legend.key.width = unit(30, "pt"),
    legend.title = element_text(margin = margin(b = 5)),
    legend.text = element_text(size = 9)
  ) +
  theme(
  panel.spacing = unit(1, "lines")
) +
  guides(
    fill = guide_colorsteps(
      title.position = "top",
      label.position = "bottom",
      barwidth = unit(14, "cm"),
      barheight = unit(0.6, "cm")
    )
  )

p
print(p)
ggsave(
  "ghana_coverage_facets_2005_2025opt_portrait.png",
  p,
  width = 14,   
  height = 24,  
  dpi = 300
)
```

```{r}
p <- ggplot(map_data) +
  geom_sf(aes(fill = Coverage_stock),
          color = "black",
          linewidth = 0.10) +
  coord_sf(expand = FALSE) +
  facet_wrap(~ year, ncol = 7) +
  scale_fill_stepsn(
    colours = c("#8c1d18", "#d7301f", "#fc8d59", "#FFD166",
                "#e5f5f9", "#99d8c9", "#2ca25f"),
    breaks  = c(0, 0.5, 0.6, 0.7, 0.8, 0.9, 0.95, 1.0),
    limits  = c(0, 1),
    labels = function(x) {
      labs <- percent(x, accuracy = 1)
      show <- x %in% c(0, 0.6, 0.8, 0.9, 1.0)  # key thresholds
      labs[!show] <- ""
      labs
    },
    na.value = "grey85",
    name = "Vaccination coverage"
  ) +
  labs(
    title = "Yellow Fever vaccination coverage by district (Ghana), 2005–2025" #,
    #subtitle = "(Optimistic)"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    strip.text = element_text(size = 10),
    legend.position = "bottom",
    legend.key.width = unit(30, "pt"),
    legend.title = element_text(margin = margin(b = 5)),
    legend.text = element_text(size = 9)
  ) +
  theme(
  panel.spacing = unit(1, "lines")
) +
  guides(
    fill = guide_colorsteps(
      title.position = "top",
      label.position = "bottom",
      barwidth = unit(14, "cm"),
      barheight = unit(0.6, "cm")
    )
  )

p
print(p)
ggsave(
  "ghana_coverage_facets_2005_2025opt_landscape.png",
  p,
  width = 24,   
  height = 14,  
  dpi = 300
)
```